---
title: "lab_02"
author: "derek willis"
date: "8/24/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings
```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
library(lubridate)
```

## Load Data

Let's keep working with West Virginia loans contained in the PPP loans data.  

```{r}
west_virginia_ppp <- read_rds("data/lab_02.rds")
```

## Examine Data

## Answer questions

Q1. Create a new column that divides the loan amounts into multiple options: under_10k, 10k_50k, 50k_100k, 100k_500k, 500k_1m, 1m_plus. How many loans are in each group?
A1. The 10k-50k group has 15,089, the most of any group.

```{r}
west_virginia_ppp %>%
  mutate(
    amount_option = case_when(
      amount < 10000 ~ 'under_10k',
      amount >= 10000 & amount < 50000 ~ '10k_50k',
      amount >= 50000 & amount < 100000 ~ '50k_100k',
      amount >= 100000 & amount < 500000 ~ '100k_500k',
      amount >= 500000 & amount < 1000000 ~ '500k_1m',
      amount >= 1000000 ~ '1m_plus'
    )
  ) %>%
  group_by(amount_option) %>%
  summarise(
    total_loans = n()
  ) %>%
  arrange(desc(total_loans))
```

Q2. Which servicing lenders have the most loans with undisbursed amounts greater than 0?
A2. Davis Trust Company tops the list with 12.

```{r}
west_virginia_ppp %>%
  filter(undisbursed_amount > 0) %>%
  group_by(servicing_lender_name, servicing_lender_state) %>%
  summarise(
    total_loans = n()
  ) %>%
  arrange(desc(total_loans))
```

Q3. Which month in 2021 had the most loans approved?
A3. March 2021, with 4,394.

```{r}
west_virginia_ppp %>%
  filter(date_approved > '2020-12-31') %>%
  mutate(Month = floor_date(date_approved, "month")) %>%
  group_by(Month) %>%
  summarise(total = n()) %>%
  arrange(desc(total))
```

Q4. How many loans went to West Virginia applicants that listed the maximum number of jobs retained (500)? Which one had the earliest approved date?
A4. There were 10 applicants that listed 500 jobs retained. J. AUSTIN HEATING & COOLING, INC, which was approved on April 8, 2020, had the earliest approved date.

```{r}
west_virginia_ppp %>%
  filter(jobs_retained == 500) %>%
  arrange(date_approved) %>%
  select(name, amount, date_approved)
```
