---
title: "lab_06"
author: "derek willis"
date: "2024-10-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

-   Tabula

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse and any other libraries you'll need
library(tidyverse)
```

## Get Our PDF

We'll be working with the [Maryland monthly voter registration update](https://elections.maryland.gov/pdf/vrar/2024_09.pdf) from the State Board of Elections. You'll want to download it to a place you'll remember (like your Downloads folder, or the labs folder in your repository). The goal is to write a couple of paragraphs that summarize the changes in voter registration in Maryland during the month of September.

## Setup Tabula

Start Tabula, then go to <http://127.0.0.1:8080/> in your browser. Click the "Browse" button and find the PDF file and click "open", and then click the "Import button" in Tabula. This will take a few seconds.

This PDF has multiple possible tables to extract. We're going to make four dataframes: 

* new registrations by party
* removals by party
* changes in registration
* current registrations by county.

You will need to draw boxes around four sections and export four CSV files that you will load into R. In Tabula, draw a box around the borders of those tables and click the "Preview & Export Extracted Data" button for each one. The export should look pretty clean, but might be missing some column headers. You'll need to include total columns.

Export each CSV (each one should be called `tabula-2024-09.csv` by default, so you should rename them for clarity) to your lab_06/data folder.

From there, you will need to read in and clean up the data, and in particular the headers. You can choose to include the headers from the PDF in your exported CSV files OR to exclude them and add them when importing. The pre_lab has an example of how to do this!

## Load and clean up the data in R

You will need to read in and clean up the data so that it can be used for analysis. By "clean" I mean the column headers must not contain spaces and they should have meaningful names, not "x1" or something similar. The current registrations dataframe *MUST* include the county name, but should not include the columns under `Changes` and `Party Affiliation From`. How you do that is up to you (and could involve something outside R), but you can use `select()` with or without the minus sign to include or exclude certain columns. You also can use the `rename` function to, well, rename columns.

```{r}
# Party registrations
new <- read_csv("data/new.csv", col_names = c("summary", "dem", "rep", "nlm", "lib", "grn", "unaf", "oth", "total", "dups"))


new
```

```{r}
# Removals
removals <- read_csv("data/removals.csv", col_names = c("summary", "dem", "rep", "nlm", "lib", "grn", "unaf", "oth", "total"))

removals
```

```{r}
# Changes
changes <- read_csv("data/changes_in_registration.csv", col_names = c("county", "address", "name", "dem", "rep", "nlm", "lib", "grn", "unaf", "other", "total"))

changes
```

```{r}
# Current registrations
current <- read_csv("data/current_registrations.csv", col_names = c("county", "dem", "rep", "nlm", "lib", "grn", "unaf", "other", "total", "conf mailing", "inactive"))

current
```

## Rinse, Repeat: September 2022 data

Do all of that again, for the [September 2022 voter registration update](https://elections.maryland.gov/pdf/vrar/2022_09.pdf), so we can compare the two. If you've done well, this is easy: just do the same steps in Tabula, export the files and use the same code to read in and clean up the data, *just remember to give these dataframes different names* so that you can compare September 2024 to September 2022.

To answer the following questions, you'll need to calculate percentages and do other calculations. Whether you do that by combining or joining dataframes or you write code to run separately on both is up to you.

```{r}
# Party registrations
new_2022 <- read_csv("data/new_2022.csv", col_names = c("summary", "dem", "rep", "grn", "lib", "wcp", "unaf", "oth", "total", "dups"))

new_2022
```

```{r}
# Removals
removals_2022 <- read_csv("data/removals_2022.csv", col_names = c("summary", "dem", "rep", "grn", "lib", "wcp", "unaf", "oth", "total"))

removals_2022
```

```{r}
# Changes
changes_2022 <- read_csv("data/changes_2022.csv", col_names=c("county", "address", "name", "dem", "rep", "grn", "lib", "wcp", "unaf", "oth", "total"))

changes_2022
```

```{r}
# Current registrations
current_2022 <- read_csv("data/current_2022.csv", col_names = c("county", "dem", "rep", "nlm", "lib", "grn", "unaf", "other", "total", "conf mailing", "inactive"))

current_2022
```

## Answer questions

Q1. Calculate the percentage of registered Democrats, Republicans and unaffiliated voters for the entire state in September 2024 and September 2022, and then calculate the percentage change for those three party categories (the formula for percentage change is easy to remember: (New-Old)/Old). How would you describe the changes in these partisan categories in a story? What's the most interesting thing about these changes that would be the lede of that story?

A1. The most interesting change in these partisan categories is the overall decrease in Democrats and increase in unaffiliated voters. That could contribute to a lede about voters straying from binary voting and becoming more neutral/indecisive in the past two years, potentially because of the Biden presidency.

```{r}
prejoin_24 <- current |>
  mutate(percent_dem24 = (dem/total * 100)) |>
  mutate(percent_rep24 = (rep/total * 100)) |>
  mutate(percent_unaf24 = (unaf/total * 100)) |>
  select (county, percent_dem24, percent_rep24, percent_unaf24)


prejoin_22 <- current_2022 |>
  mutate(percent_dem22 = (dem/total * 100)) |>
  mutate(percent_rep22 = (rep/total * 100)) |>
  mutate(percent_unaf22 = (unaf/total * 100)) |>
  select (county, percent_dem22, percent_rep22, percent_unaf22)

inner_join(prejoin_22, prejoin_24) |>
  mutate (percent_demchange = (percent_dem24-percent_dem22)) |>
  mutate (percent_repchange = (percent_rep24-percent_rep22)) |>
  mutate (percent_unafchange = (percent_unaf24-percent_unaf22))

```

Q2. Among the Maryland jurisdictions with the most registered voters, which one had the biggest percentage increase in unaffiliated voters between 2022 and 2024? How did that county vote for governor in 2022 and for president in 2020? Contrast the partisan percentages of that county with the county that had the eighth-largest percentage increase in unaffiliated voters. You should display information from the data that shows the counties with the largest percentage increase first.

A2. Prince George's county had the biggest percent change in 2022. They voted for Wes Moore for governor and Joe Biden in 2020. Prince George's lost more Democrats than Baltimore County and gained more Republicans and unaffiliated voters.

```{r}

inner_join(prejoin_22, prejoin_24) |>
  mutate (percent_demchange = (percent_dem24-percent_dem22)) |>
  mutate (percent_repchange = (percent_rep24-percent_rep22)) |>
  mutate (percent_unafchange = (percent_unaf24-percent_unaf22)) |>
  select (county, percent_demchange, percent_repchange, percent_unafchange) |>
  arrange(desc(percent_unafchange))


```

Q3. Do the party percentages of new registrants (% Dem, % Rep, % Unaffiliated) from September 2024 and September 2022 match the overall percentage of partisan registration in the state at those times?

In other words, did new registrants mirror the overall partisan composition of the electorate, or were they different? Do any of the methods of registration vary significantly from the overall partisan pattern? Which method is worthy of further exploration?

A3. In 2024, there was higher amounts of democratic registration with more republican voters; in 2022, the numbers seem inconsistent, with lots of democratic registration and split partisan registration. It seems like republicans tend to opt for in person voting while democrats prefer absentee and mail in ballots. I wonder if the lack of republican mail-in voting is due to Trump's skepticism about voter authenticity.

```{r}
new |>
  mutate(percent_dem24 = (dem/total * 100)) |>
  mutate(percent_rep24 = (rep/total * 100)) |>
  mutate(percent_unaf24 = (unaf/total * 100)) |>
  select(summary, percent_dem24, percent_rep24, percent_unaf24)

new_2022 |>
  mutate(percent_dem22 = (dem/total * 100)) |>
  mutate(percent_rep22 = (rep/total * 100)) |>
  mutate(percent_unaf22 = (unaf/total * 100)) |>
  select(summary, percent_dem22, percent_rep22, percent_unaf22)

current |>
  mutate(percent_dem24 = (dem/total * 100)) |>
  mutate(percent_rep24 = (rep/total * 100)) |>
  mutate(percent_unaf24 = (unaf/total * 100)) |>
  select(percent_dem24, percent_rep24, percent_unaf24) 

current_2022 |>
  mutate(percent_dem22 = (dem/total * 100)) |>
  mutate(percent_rep22 = (rep/total * 100)) |>
  mutate(percent_unaf22 = (unaf/total * 100)) |>
  select(percent_dem22, percent_rep22, percent_unaf22)

```

Q4. Choose your own adventure! Using both the 2024 and 2022 data for one of the dataframes (changes, removals or new registrations), ask and answer a question that makes at least one comparison and requires code to answer. That could involve calculating a percentage, percentage change or the raw difference between two columns.

A4. I'm looking to compare the amount of republican voters that use mail-in ballots overtime and if that number hs increased.

```{r}
new |>
  mutate(percent_dem24 = (dem/total * 100)) |>
  mutate(percent_rep24 = (rep/total * 100)) |>
  select(summary, percent_dem24, percent_rep24)

new_2022 |>
  mutate(percent_dem22 = (dem/total * 100)) |>
  mutate(percent_rep22 = (rep/total * 100)) |>
  select(summary, percent_dem22, percent_rep22)
```

Q5. What's the best story idea or question you've seen as a result of the work you've done in this lab?

A5. Have Republicans become less trusting of mail-in ballots since Trump's presidency? Compare data from pre-Trump to post-Trump.
